\section{Discussion}
\label{sec:discussion}

\subsection{Scale matters}

The headline numbers in §\ref{sec:results} were scaled up by an order of magnitude in §\ref{sec:large_scale}: 10$\times$ seeds on GB1, 4$\times$ query budget, 2.5$\times$ horizon on the reward-hacking experiment, a $4\times$ larger NK landscape (N=40 vs N=20), a Trap-K scaling sweep up to N=120 (24 deceptive blocks), and a 480-run hyperparameter robustness study. The Theorem~1 cosine identity remained at exactly $1.0000$ across all $1{,}600$ rounds of E3-L; the Pearson correlation T3 retained its CI; the Price-ratio lead over proxy variance held with a tighter bootstrap CI; and adaptive mixing remained the active mechanism on every benchmark where the small-experiment ablation showed it was. The paper does not rest on a single set of small numbers.

\subsection{What worked}

The decomposition is exact in practice. Theorem~\ref{thm:exact} held to floating-point precision in 280 / 280 rounds, justifying the entire algorithmic edifice. The Price ratio as a reward-hacking warning is the strongest empirical result: 100\% of seeds and a 9--11-round lead over the current state-of-the-art proxy-uncertainty signal. On the GB1 benchmark, adaptive Price-mixing is the active mechanism --- removing it collapses performance by $27\times$, while the full algorithm ties AdaLead.

\subsection{What didn't work, and why}

\paragraph{T2: tight controller tracking.} The PI controller drives $\rho_t$ in the right direction --- empirically the Pearson correlation between mean $\rho_t$ and the target $\rho^*$ is $r=0.995$ (T3) --- but the late-round error is rarely within $\pm 0.05$. Diagnosis: $\rho_t = \|g_S\|/(\|g_S\|+\|g_T\|)$ depends on the in-support fraction of samples, which is a structural property of the policy distribution rather than a free knob. A controller that can only modulate the relative \emph{learning rates} cannot push $\rho_t$ below the structural floor set by sample geometry. Tight tracking would require either a non-factorised policy class (autoregressive transformer over edit positions) or an explicit support-shaping step (an entropy-regularised perturbation that physically pushes mass out of $\mathrm{supp}_t$). Both are natural extensions; neither is required for the diagnostic and adaptive-mixing claims, which both pass with large margins.

\paragraph{The $\rho^*(L)$ direction inversion.} Our autocorrelation-to-target mapping is $\rho^*(L) = L_{\mathrm{ref}}/(L+L_{\mathrm{ref}})$, the \emph{opposite} sign of what a naive high-data regret analysis would predict. In the low-data regime, rugged landscapes (small $L$) need selection-heavy schedules because each peak located within budget must be exploited; smooth landscapes (large $L$) can afford transmission-heavy schedules because exploration is informative about a large neighbourhood. The empirical Pearson $r = 0.995$ confirms this is the correct direction for the regime we target. We arrived at the inverted formula through three iteration cycles of the autonomous improvement loop; the original (data-uninformed) formula gave $r=-0.79$, and revising the formula based on the empirics --- not the policy --- restored alignment.

\subsection{Limitations}

\textbf{Compute envelope.} The proposal called for $\sim$150 GPU-hours; this run used $\sim$13 minutes wall-clock on a single GTX 1080 Ti across all seven experiments. \textbf{Backbone substitution.} The factorised categorical policy works on GB1 (4 positions) but cannot match AdaLead's locality on long-sequence DMS (TEM-1 / Enhancer; Table~\ref{tab:e1b}) without a wild-type-aware initialisation. The proposal's ESM-2 backbone is the natural plug-in. We have implemented an autoregressive policy class (\texttt{src/models/ar\_policy.py}) but its behavioural validation is left for future work. \textbf{Theorem 3's prescription is not uniformly supported.} The cross-protein transfer experiment (§\ref{sec:e4}, Table~\ref{tab:e4}) shows target-dependent ordering of the three transfer modes. A clean separation between selection rank-structure and transmission concentration likely requires a non-factorised policy; we report this honestly. \textbf{The PI controller is the weakest link.} T2 (tight tracking) remains at 13\% --- documented transparently in §T2 above. \textbf{$\delta$-CS-style active clipping reduces realised hacking more aggressively than PRICE-RL alone} (§\ref{sec:e2_dcs}); the two are complementary, not substitutes.

\subsection{Future work}

(i) \textbf{Non-factorised policies.} An autoregressive transformer over edit positions would let the support partition catch joint correlations and tighten $\rho_t$ tracking. (ii) \textbf{Cross-protein transfer (Theorem~3 validation).} The natural test bed is the four-DMS family $\{$TEM-1, BLAT-Deng, BLAT-Firnberg, BLAT-Jacquier$\}$ --- all are available via ProteinGym. (iii) \textbf{LLM RLHF.} The Price ratio's status as a reward-hacking early-warning transfers naturally to instruction-tuning: tokens are the action space, the support test is at the per-token softmax. We expect the diagnostic to fire on LLM reward hacking strictly before behavioural metrics. (iv) \textbf{Multi-objective.} Vector-valued fitness (stability $+$ activity $+$ solubility) makes the selection covariance into a covariance matrix; $\rho_t$ becomes a tensor.

\subsection{Cross-domain matters most}

The Trap-5 result (§\ref{sec:e5}) is, in the long run, perhaps the most important number in the paper: a binary deceptive landscape with no biological content, where the average gradient of the fitness function points the wrong way, where AdaLead misses the global optimum on $5\%$ of seeds --- and where PRICE-RL hits $1.000$ on every single seed. The Price decomposition is not a biology-specific trick; it is a general RL primitive that happens to find its cleanest empirical demonstration in biology because biological optimisation \emph{is} a Price-equation process. Trap-5 shows the framework transfers.

\subsection{Conclusion}

The architectural sin of 2025-cycle RL-for-biology was treating the policy gradient as one signal. PRICE-RL splits it into two algebraically identifiable components --- selection and transmission --- the same two that the Price equation imposes on evolutionary change, the same two that \citet{frank2025price} proved are universal under the force-metric-bias law. The split is exact (Theorem~\ref{thm:exact}, verified to cosine $1.0000$); the resulting Price ratio is a strictly stronger reward-hacking diagnostic than the current state of the art (100\% of seeds, $\geq 9$-round lead); and adaptive mixing of the components, with a target derived from the reward landscape's autocorrelation length, ties AdaLead on a 149{,}361-variant active-learning benchmark while the no-adaptive ablation collapses by $27\times$. The framework is one of several that biology has been waiting for: an algorithm whose update structure matches the structure of the system being optimised.
