\section{Experimental setup}
\label{sec:experiments}

We test three pre-registered hypotheses corresponding to the three theorems, and we extend each of them into a matching scaled study in §\ref{sec:large_scale} so that no headline conclusion rests on a single small experiment. \textbf{H1 (low-data SOTA, partial):} PRICE-RL matches state-of-the-art active-learning baselines on the 149{,}361-variant GB1 four-mutation landscape under a 5-round / 100-query budget, and the adaptive-mixing ablation collapses to near-random, isolating the controller as the active mechanism rather than the decomposition alone. \textbf{H2 (reward-hacking diagnostic):} on a deliberately weak proxy surrogate, $\rho_t$ saturates earlier than proxy ensemble variance, the signal used by $\delta$-CS \citep{kim2025deltacs}; the relevant comparison is per-round-lead time rather than realised reward gap, because the gap is a downstream consequence of the gradient that the diagnostic flags. \textbf{H3 (theory--empirics match):} on NK landscapes \citep{kauffman1989nk} swept across $K\in\{0,1,3,5,10,15,19\}$, the empirical mean $\rho_t$ tracks the analytic regret-optimal $\rho^*_t$ as $K$ varies, with the autocorrelation length serving as the only externally fitted quantity. The three hypotheses are deliberately disjoint: H1 measures end-to-end performance, H2 measures the diagnostic, and H3 measures the alignment between empirics and the regret-by-autocorrelation theory that motivates the controller target.

\paragraph{Datasets.} We use \textbf{GB1 four-mutation} \citep{wu2016adaptation} via the FLIP repository \citep{dallago2021flip} (149{,}361 four-position combinatorial variants of the immunoglobulin-binding protein G domain); \textbf{GFP} \citep{sarkisyan2016local}, \textbf{TEM-1 $\beta$-lactamase} \citep{stiffler2015evolvability}, and \textbf{AAV capsid} \citep{bryant2021deep} via ProteinGym v0.1 \citep{notin2023proteingym}; the \textbf{BCL11A enhancer} \citep{kircher2019saturation} via MaveDB \citep{esposito2019mavedb}; and synthetic \textbf{NK landscapes} with $N\in\{20,40\}$, alphabet $A=4$, $K\in\{0,1,3,5,10,15,19\}$, with deterministic Blake2b-keyed per-locus contributions giving exact reproducibility across machines. The analytic autocorrelation length is $L_{\mathrm{NK}}(K) = N/(K+1)$ \citep{stadler1999random}. The deceptive \textbf{Trap-K} family is included as a non-biological cross-domain stress test.

\paragraph{Baselines.} We compare against \textbf{Random} (uniform), \textbf{AdaLead} \citep{sinai2020adalead} (greedy evolutionary baseline with edit rate $\kappa = 0.05$ and 8 parents), \textbf{PEX} \citep{ren2022proximal} (proximal exploration with edit-distance regularisation, radius~3), and a faithful \textbf{GFlowNet-AL} reimplementation (flow-matching factorised categorical, soft-Q-style training on labelled history, per \citealp{jain2022biological}) at the larger budgets where its training data is sufficient. Two PRICE-RL ablations isolate where the gain comes from: \textbf{PRICE-RL ($\alpha_S=\alpha_T$)} fixes the mixing, removing only the controller, and the \textbf{open-loop schedule} sets $\alpha_S \propto \rho^*_t$ without feedback. A \textbf{random-support} ablation that uses a randomly-coined partition with the same in-support fraction confirms that the partition itself, not just the mixing schedule, carries the signal.

\paragraph{Protocols.} \textbf{E1 (GB1)} runs 5 rounds at $B = 100$ over 5 seeds; the primary metric is top-$1\%$ recall, the discriminative metric in this regime because best fitness saturates almost immediately at the trivial threshold, and the secondary metric is best fitness reached. We compare methods with a 10{,}000-resample paired bootstrap on the difference of means and an explicit non-inferiority test within $1$ std for the AdaLead comparison, so that ties are reported as ties rather than as wins for whichever method happens to land slightly higher. \textbf{E2 (reward hacking)} uses NK($N=20, K=10$) with a 200-label MLP surrogate as the proxy reward; PRICE-RL runs for 12 rounds against the surrogate while the true fitness is logged but never seen by the policy, and the proxy variance signal is computed from a 3-model ensemble trained on the same labels. We measure both the lead-time of the $\rho_t \geq 0.9$ alarm against the proxy-variance crossing and the realised proxy-vs-true reward gap, because the two signals capture different aspects of hacking (early diagnosis vs.\ end-of-campaign harm). \textbf{E3 (NK sweep)} runs seven $K$ values $\times$ five seeds $\times$ 8 rounds at $B=64$; we log per-round $\rho_t$, $\rho^*_t$, $\hat L_t$, and the cosine of $g_S+g_T$ against the pooled REINFORCE gradient as a continuous Theorem~\ref{thm:exact} self-check. The matching scaled studies (E1-L at $4{,}000$ queries, E1-mega at $8{,}000$ queries, E2-L at $30$ rounds and $16$ seeds, E3-L at $N=40$ over $1{,}600$ rounds, E5-L over a $\{30,60,120\}\times\{3,5,7,10\}$ Trap-K grid, and E6 over $480$ hyperparameter configurations) extend the same protocols to budgets $4$--$16\times$ larger; full details are in §\ref{sec:large_scale} and Appendix~\ref{app:reproducibility}, and the headline conclusions hold in every scaled study.
