\section{Introduction}
\label{sec:introduction}

The 2025 NeurIPS / ICML cycle established a dominant template for reinforcement-learning-driven biological sequence design: take a pretrained generative model (a protein language model, a discrete diffusion model, or a DNA language model); fine-tune with PPO, DPO, GRPO, or KTO against a fitness proxy; add a naturalness regulariser. Recent representative methods include $\delta$-Conservative Search \citep{kim2025deltacs}, $\mu$Protein and $\mu$Search \citep{sun2025muprotein}, the Caltech / Yue-lab steering work of \citet{yang2025steering}, GLID$^2$E \citep{cao2025glid2e}, BioReason \citep{bioreason2025}, and the Bayesian active-learning toolkit ALDE \citep{yang2025alde}. They all share a single architectural choice: an \emph{undifferentiated} policy gradient against a scalar reward.

This pooling is the central pathology of RL-for-biology in 2025. When the proxy reward becomes unreliable on out-of-distribution sequences --- the regime in which a designed variant departs from the labelled support of the surrogate --- the policy hacks the proxy. Existing mitigations such as $\delta$-CS clip the exploration radius using proxy variance, treating the symptom (mis-calibrated rewards) rather than the cause (a gradient estimator that cannot tell us whether the policy is exploiting known regions or shifting into new ones). The policy never reports back, per round, on which of those two operations its update is actually performing.

The cause is structural. The standard policy gradient
\(
\nabla_\theta J(\theta) = \mathbb{E}_{x \sim \pi_\theta}\!\left[ R(x)\,\nabla_\theta \log \pi_\theta(x) \right]
\)
pools two effects into one estimator: it up-weights trajectories the policy already supports (the analogue of evolutionary \emph{selection}), and it shifts the policy's support to new regions (the analogue of \emph{transmission}, i.e.\ mutation and inheritance). \citet{price1970selection} decomposed any change in mean trait value across an evolving population into precisely these two terms, exactly. \citet{frank2025price} recently proved that the algebraic form $\Delta\theta = M f + b + \xi$ --- the force--metric--bias law --- subsumes natural selection, Bayesian updating, SGD, and policy gradients as algebraically identical objects, with the Price decomposition lifting to all of them. Crucially, that paper made no algorithmic claim. It established that the decomposition was \emph{available}; it did not show what one could do with it. We do.

The starting point is a simple identity: every Monte-Carlo policy gradient sample $x$ either lies in the previous policy's support or it does not, and the indicator $\mathbb{1}_{\mathrm{supp}_t}(x) + \mathbb{1}_{\overline{\mathrm{supp}}_t}(x) = 1$ partitions the gradient sample-wise. Multiplying through by the per-sample policy-gradient term gives $\hat g_S + \hat g_T = \hat g_{\mathrm{pool}}$ exactly --- not in expectation, but as random variables on every batch. The two halves can therefore be measured, weighted, and gated independently without affecting the estimator's unbiasedness. A single scalar $\rho_t = \|g_S\|/(\|g_S\|+\|g_T\|)$ summarises the gradient's selection--transmission balance on a unit interval, and the same scalar serves both as a controller signal and as a diagnostic.

The diagnostic role is the cleanest part of the story. Reward hacking is structurally an event in which a policy refines an artefact within its existing support: the proxy is largest on a sequence the policy already places mass on, and the gradient updates push that mass higher rather than spreading it. That picture has a direct fingerprint in the Price ratio: $\rho_t$ saturates near $1$ several rounds before the realised proxy-vs-true gap or the proxy ensemble variance has time to reach its alarm. We use this property as the foundation of a per-round, gradient-level reward-hacking warning. The controller role is more delicate: the right balance between the two components depends on the reward landscape itself, and we tie the target $\rho^*_t$ to an autocorrelation-length estimate $\hat L_t$ via $\rho^*(\hat L) = L_{\mathrm{ref}}/(\hat L + L_{\mathrm{ref}})$. Smooth landscapes admit transmission-heavy schedules because exploration is cheap; rugged landscapes require selection-heavy schedules because each peak located within budget must be exploited. A PI controller drives the empirical $\rho_t$ towards $\rho^*_t$ by modulating the relative learning rates $\alpha_S, \alpha_T$ while holding their geometric mean fixed.

\paragraph{Contributions.} (i) A new RL primitive: estimate the selection and transmission components of $\nabla J$ separately on policy, partitioned by membership in the previous policy's support, with a sample-wise exact identity that requires only the existence of $\nabla_\theta \log \pi_\theta$ and so applies to factorised \emph{and} autoregressive parameterisations. (ii) An adaptive controller that drives the empirical Price ratio $\rho_t$ towards a target $\rho^*_t$ derived from the reward landscape's autocorrelation length, with theoretical regret guarantees and an empirical Pearson $r=0.995$ between $\rho_t$ and the regret-optimal target across an NK ruggedness sweep. (iii) A new diagnostic: $\rho_t$ itself, an early-warning signal for reward hacking that fires gradient-side rather than behaviour-side and, in our experiments, beats the proxy-variance signal of $\delta$-CS by $9$--$18$ rounds on $100\%$ of seeds. (iv) Theorem~1 (exactness for factorised categorical policies, with an autoregressive corollary), empirically validated to cosine $1.0000$ on every round of every NK seed in the $1{,}600$-round large sweep. (v) Empirical results on $149{,}361$-variant GB1 \citep{wu2016adaptation}, four ProteinGym DMS landscapes, the binary deceptive Trap-K family, and a $480$-config robustness study. (vi) Public release of the code under MIT licence.

\paragraph{Where this is and is not novel.} It is \emph{not} novel to use RL for biological sequence design \citep{angermueller2020dyna,jain2022biological,lee2024latprotrl,kirjner2024smoothed}; it is \emph{not} novel to be ruggedness-aware \citep{towers2024ruggedness,sun2025muprotein}; it is \emph{not} novel to relate evolution and learning at a verbal level \citep{salimans2017evolution}. What \emph{is} novel is the formal operationalisation of \citeauthor{frank2025price}'s 2025 unification as an algorithmic primitive, the per-round Price-ratio diagnostic, and the regret-by-autocorrelation framing of the controller target. The paper does not propose a better protein optimiser per se; it proposes that the algebraic structure of the optimiser was wrong, and shows that biology is the domain where the rewriting pays off most cleanly because biological optimisation already \emph{is} a Price-equation process.
