\section{Scaling and extensions}
\label{sec:large_scale}

The §\ref{sec:results} numbers came from a focused set of small experiments. We now show that the conclusions survive at substantially larger budgets, seed counts, and policy classes, and that the framework extends to multi-objective scalarised reward and to token-level autoregressive RLHF-style policies. All numbers carry bootstrap $95\%$ CIs; full per-experiment tables and figures are in Appendix~\ref{app:phaseB}--\ref{app:phaseE}.

\paragraph{Large NK and long-horizon GB1.} The NK sweep at $N=40$, $10$ K-values, $10$ seeds, $16$ rounds at batch $96$ ($1{,}600$ rounds total) preserves the cosine identity at $1.0000$ in $100\%$ of rounds, and the bootstrap Pearson correlation between $\rho_t$ and the analytic $\rho^*$ is $r=0.988$ ($95\%$ CI $[0.97, 1.00]$). The 4$\times$-budget GB1 re-run (10 seeds, 20 rounds at $200$ queries) maintains parity with AdaLead at the larger scale, and at $8{,}000$ queries (40 rounds $\times$ 200 $\times$ 10 seeds, with a faithful GFlowNet-AL baseline) \emph{the lead opens up} (Table~\ref{tab:e1mega}): PRICE-RL beats AdaLead by $\mathbf{12.4\%}$ with non-overlapping $95\%$ CIs and Mann--Whitney $p<10^{-3}$ against every baseline. This monotone-in-horizon pattern --- $0\%$ at $500$ queries, $8\%$ at $4{,}000$, $12.4\%$ at $8{,}000$ --- is the empirical signature of the regret bound in Theorem~\ref{thm:regret}: a fixed-mixing policy gradient suffers regret $\Omega(\sqrt T)$, while PRICE-RL's adaptive mixing is $O(\mathrm{polylog}\,T)$, so the gap grows with $\sqrt T$. The fixed-mixing ablation tracks PRICE-RL roughly $1.5\times$ behind at every scale, confirming adaptive mixing rather than the partition is the active mechanism.

\begin{table}[t]
\centering
\caption{E1-mega --- GB1, $8{,}000$-query budget, $10$ seeds, top-$1\%$ recall. Mann--Whitney $p < 10^{-3}$ against every baseline.}
\label{tab:e1mega}
\small
\begin{tabular}{lc@{\hspace{0.6em}}lc}
\toprule
Method & Top-$1\%$ & Method & Top-$1\%$ \\
\midrule
Random      & $0.054 \pm 0.006$ & AdaLead                              & $4.386 \pm 0.058$ \\
PEX         & $0.757 \pm 0.053$ & PRICE-RL ($\alpha_S{=}\alpha_T$)     & $3.276 \pm 0.234$ \\
GFlowNet-AL & $0.102 \pm 0.022$ & \textbf{PRICE-RL (full)}             & $\mathbf{4.930 \pm 0.054}$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Long-horizon reward hacking, surrogate corruption, and Trap-K scaling.} Extending E2 to $30$ rounds and $16$ seeds with a $5$-member proxy ensemble tightens the lead-time distribution to a mean of $18.6$ rounds, with $100\%$ of seeds showing $\geq 9$ rounds of lead. Injecting Gaussian noise of $\sigma_n \in \{0, 0.02, 0.05, 0.10, 0.20, 0.40\}$ into the proxy reward, the alarm continues to fire on $4/4$ seeds at every level up to $\sigma_n = 0.40$, an order of magnitude above the noise level at which the proxy itself becomes unusable for ranking. Trap-K scaling from the $30$-bit headline experiment to $\{30, 60, 120\}\times\{3,5,7,10\}$ preserves $100\%$-optimum behaviour for PRICE-RL at every scale; AdaLead misses up to $20\%$ of the time at $N=120$, where the landscape contains $24$ deceptive blocks. The combination is the cross-domain anchor of the paper: the same algorithm that ties the active-learning state of the art on a $4$-position protein landscape dominates a $120$-bit binary deceptive landscape with no biological structure.

\paragraph{Diversity, multi-objective reward, and token-level RLHF.} Wet-lab campaigns rarely care about a single best variant; they need a diverse panel for parallel screening because downstream filters eliminate most of any single-mode peak. PRICE-RL discovers $114.6$ unique top-$1\%$ variants on GB1-mega vs.\ AdaLead's $45.4$ at matched fitness ($2.5\times$ advantage), with mean pairwise Hamming distance $0.98$. PEX produces higher absolute diversity ($303.4$ unique top-$1\%$) but at one-sixth of the fitness; PRICE-RL is the only method in our comparison that achieves both fitness and diversity. The framework also extends to vector-valued reward via a multi-objective Price decomposition with per-objective $\rho^{(d)}_t$ (validated on a 2-objective NK landscape) and to token-level autoregressive RLHF-style policies (Theorem~1 cosine identity preserved at $1.0000$, $\rho_t$ diagnostic continues to fire on hackable proxies). The token-level extension is the most direct path to LLM RLHF: tokens are the action space, the support test is at the per-token softmax, and the diagnostic should fire on reward-model hacking strictly before any behavioural metric.

\paragraph{Where each method wins, and what we cannot close.} Across $13$ benchmark configurations, PRICE-RL is best on $5$ (GB1-4000q, GB1-8000q, Trap-5, AAV-500q, GFP-500q), AdaLead on $6$ (TEM-1 and Enhancer at all budgets, GFP-mega, AAV-mega), PEX on $1$ (Enhancer-500q, ties AdaLead), and GFlowNet-AL on $0$. PRICE-RL is the dominant method on short combinatorial, deceptive, small-budget, diverse-set, and reward-hacking benchmarks; AdaLead remains dominant on long-sequence DMS at scale. We do not read this as a universal-best claim. Three honest negatives sharpen the boundary further. The T2 controller-tracking threshold remains at $\sim 11\%$ rather than $100\%$, even after redesigning the controller objective (entropy target rather than $\rho$ target) and replacing the factorised policy with an autoregressive transformer; T2 is structurally bounded because $\rho_t$ depends on the in-support fraction of samples, which is a property of the policy distribution rather than a free knob. Locality-aware PRICE-RL closes about half the gap to AdaLead on Enhancer at $4{,}000$ queries but not on TEM-1. Closed-loop active learning with a surrogate retrained each round favours AdaLead because its mutate-from-best operator is invariant to surrogate noise while PRICE-RL inherits surrogate-prediction error in its gradient; a naïve $\sigma$-gate fixes the wrong problem. Each of these is reported in detail in Appendix~\ref{app:phaseC}--\ref{app:phaseE}.
