\section{Related work}
\label{sec:related}

\paragraph{RL for biological sequence design.} The 2025 main-track methods include $\delta$-Conservative Search \citep{kim2025deltacs}, $\mu$Protein and $\mu$Search \citep{sun2025muprotein}, the Caltech / Yue-lab steering work \citep{yang2025steering}, ALDE \citep{yang2025alde}, GLID$^2$E \citep{cao2025glid2e}, and the DNA-LLM RLVR setup of BioReason \citep{bioreason2025}. Earlier diversity-aware approaches include GFlowNets \citep{jain2022biological}, AdaLead \citep{sinai2020adalead}, PEX \citep{ren2022proximal}, and DyNA-PPO \citep{angermueller2020dyna}; latent-space RL for rugged proteins is treated by LatProtRL \citep{lee2024latprotrl} and smoothed-landscape RL by \citet{kirjner2024smoothed}. \emph{None} of these methods decomposes the policy gradient into selection and transmission components, and all inherit the undifferentiated-gradient pathology of §\ref{sec:introduction}; PRICE-RL's decomposition is orthogonal rather than competitive, and could in principle wrap any of them.

\paragraph{Price equation in evolution and learning theory.} \citet{price1970selection} gave the exact decomposition $\Delta\bar f = \mathrm{Cov}(w,f)/\bar w + \mathbb{E}[w \cdot \Delta f]/\bar w$, used as an analytical tool in evolutionary biology for fifty years \citep{frank2012natural}. Its connection to learning algorithms remained metaphorical until \citet{frank2025price} proved that $\Delta\theta = M f + b + \xi$ specialises to natural selection, Bayesian updating, Newton's method, SGD, Adam, and the policy gradient as instances of the same force--metric--bias law. That paper supplied the unifying algebra; PRICE-RL is, to our knowledge, the first algorithmic instantiation that makes the decomposition an active component of an update rule rather than a re-description of one.

\paragraph{Reward hacking and structured policy gradients.} \citet{skalse2022defining} formalised reward hacking; \citet{laidlaw2025correlated} introduced occupancy-measure mitigations. Both are post-hoc tests on aggregate behaviour. The proxy-uncertainty signal of $\delta$-CS \citep{kim2025deltacs} is the closest comparator and operates round-by-round, but it lives at the predictor side rather than the gradient side, asking \emph{is the surrogate reliable on this batch?} The Price ratio asks \emph{is the gradient consuming new sequence space or refining what the policy already supports?}, which is the structurally earlier question. Two-timescale RL \citep{konda2000actor,karimi2022two} uses different rates for distinct function approximators (policy vs.\ value); PRICE-RL uses different rates for decomposed components of \emph{the same} gradient, partitioned by support membership rather than by network identity.
